Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-14 paper.tex
Some corrections following @SilviaOlabarriaga's comments.

\begin{abstract}
	\note{sil}{i find this a dry start. what about an intro sentence?
		Workflow engines are critical for the efficient and transparent
		exploitation of distributed infrastructures in the ecosystem of tools
		and services offered by science gateways.}
  We investigate six software architectures commonly used to integrate
  workflow engines into science gateways. In \emph{tight integration},
  the workflow engine shares software components with the science
  gateway. In \emph{service invocation}, the engine is isolated and
  invoked through a specific software interface. In \emph{task
    encapsulation}, the engine is wrapped as a computing task executed
  on the infrastructure. In the \emph{pool model}, the engine is
  bundled in an agent that connects to a central pool to fetch and
  execute workflows. In \emph{nested workflows}, the engine is
  integrated as a child process of another engine. In \emph{workflow
    import}, the engine is integrated through workflow language
  conversion. We describe and evaluate these architectures with
  metrics for assessment of integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. 
  \note{sil}{what are the insights for wf engine developers?. this is not clear in the discussion and conclusions. i think the focus lies more on the integration and not on one or the other} Tight
  integration and task encapsulation are the easiest to integrate and
  the most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}{Tristan}{Added a description of the
    architectures, following @pbellec's suggestion.}
\note{sil{what about closing with this sentence from above:
		These results provide insights for
		science gateway architects and workflow engine developers. }
\end{abstract}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties introduced in Section~\ref{sec:intro_sg}. Criteria break down to specific metrics where
\emph{lower value indicates better performance}. For each criterion, a
global score is computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score for a criterion are comparable, so that they can be combined meaningfully. The criteria and metrics are
explained hereafter.

\note{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}

@@

\note{Sil}{i was wondering whether it would be more clear to present the metrics like this:
\item \texttt{I$_1$}: total number of components, and

now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface
	}
Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Number of specific components
involved in workflow executions (\texttt{R$_1$}). The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Number of specific interactions
involved in workflow executions (\texttt{R$_2$}).
\note{Sil}{add here an example of an interaction that is not, for example, data transfers?}
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\item New engine \texttt{E$_1$}: number of interactions or
  components that need to be modified to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
  \note{Si}{new interface = new interaction? maybe better to use one term only here. Maybe also refer to how this shows on figure 2}
\item Engine version upgrade \texttt{E$_2$}: number of interactions or
  components modified to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated

@@

  the infrastructure, etc.
  \note{Silv}{it seems to me that this metric cannot be calculated in a straighforward manner from the diagrams.. is that correct? if it can, give an example of how this is seen in the diagram}
\item New workflow \texttt{E$_3$}: number of interactions required
to integrate a new workflow in the architecture. Adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions, assuming
  that the engine type and version required to execute this workflow
  are already available.
  \note{sil}{again here it would be good to indicate how this is seen in the diagram}

@@

\item Multiple engine instances (\texttt{S$_1$}): measures the possibility \note{sil}{are we measuring possibility or potential? i think the second, because it is hypothetical. possibility brings me the idea of it is possible or not - which would be zero to almost all these features} of having

@@

  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number

@@

  intervention (e.g., two-factor authentication \note{sil}{what is this?}), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 means that a new engine

@@

\item Distributed engines (\texttt{S$_2$}): measures the possibility of distributing

@@

  engine. \texttt{S$_2$}=0 means that distributed engines are possible in

@@

\item Task scheduling (\texttt{S$_3$}): measures ??? \note{sil}{need to be more specific about what is being called task scheduling here. also explain why it is difficult, or dont say anything if it is common knowledge}. Task scheduling typically depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complicate the task
  scheduling problem by introducing additional software layers or
  creating tasks with specific characteristics. \texttt{S$_3$}=0 means that
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.

@@

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that the feature is available, and
  \texttt{O$_1$}=1 otherwise.
  \note{si}{why here we do not have the "possible to implement" situation?}

@@

\paragraph{Scalability} The service architecture in principle supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17. A basic
load-balancing mechanism is available that sends new workflow
executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
 instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1). 
\note{sil}{is this score for VIP? i think that the score depends on the implementation of the science gateway, which could adopt some elastic mechanism within. so we should not say s_2=1 for the architecture}
The scheduling of tasks on the
infrastructure is as complex as in any other architecture since the
workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).
\note{sil}{see comment in section 3.1}

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
\note{sil}{i have the impression that the comments below are about robustness, and not so much about scalability. i also think that the level of detail here is much larger than in other 3.X sections. finally, i wonder whether what is explained here (which i must admit i could not follow completely in the scope of this evaluation or architectures) is valid only for this architecture in particular. i have the impression that this argument is more general }
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. \closednote{Naj}{The lay reader wonders why?  an
  example like the one offered for interdependencies can
  help}{Tristan}{Done, thanks.} Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen. \closednote{Naj}{so is this the solution to
  fixing the above mentioned limitations with
  scalability?}{Tristan}{Unfortunately this is not a solution. We just
  say that this situation is unlikely to happen which is true.}

@@

\paragraph{Scalability} As in the service architecture, adding a new
workflow instance requires manual configuration in the science gateway
(instance of a parent engine), or in the parent engine (instance of a
child engine) (\texttt{S$_1$=1}). Similarly, elastic engines are
difficult to achieve. Distributed engines are
possible, through meta-workflows (\texttt{S$_2$=0}). Task scheduling
is more complex than in other architectures though, due to the fact
that workflow execution is split in different engines (\texttt{S$_3$=1}).

\paragraph{Specific features}
Meta-workflows are possible, which is one of the main interest of this
architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
science gateway cannot directly access fine-grained information in the
sub-workflow (\texttt{O$_2$=1}).

@@

Finally, workflow engines are only one of the many aspects involved in
the design of a science gateway. Other properties definitely influence
the design of a software architecture, for instance collaborative
features, data visualization, search, authentication, accounting and
so on.
\note{sil}{do you want to refer to the science gateway canvas paper here?
	@inproceedings{shahand:2015ab,
		Address = {Portland, OR, USA},
		Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
		Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
		Series = {SCREAM'15},
		Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
		Year = {2015},
	}
\section{Conclusion}

@@

This work is in the scope of the LABEX PRIMES (ANR-11- LABX-0063) of
Universit\'e de Lyon, within the program ``Investissements d’Avenir''
(ANR-11-IDEX-0007) operated by the French National Research Agency
(ANR). This work also falls into the scope of the scientific topics of
the French National Grid Institute (IdG). We thank the site
administrators of the European Grid Initiative and the GGUS support
for their help related to the VIP platform. The CBRAIN team is grateful for the
computing cycles, storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from CANARIE
 (\url{http://www.canarie.ca/}). The integration between PSOM and CBRAIN was supported by a Brain 
Canada Platform Support Grant, as well as the Canadian Consortium on
Neurodegeneration in Aging (CCNA), through a grant from the Canadian Institute
of Health Research and funding from several partners. This work has been made possible with the support of the Irving Ludmer Family Foundation and
the Ludmer Centre for Neuroinformatics and Mental Health. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not sure if
  Canarie should be added. What about Calcul QC?}{Marc-e}{CQ, no. CANARIE, I'd
  say yes since they funded a major chunk of CBRAIN's development.}

\note{sil}{should we then also thank the dutch crowd?
	 We also acknowledge the Dutch national e-Infrastructure with the support of SURF Cooperative,  the Dutch national program COMMIT/ and the High Performance Computing and Networking (HPCN) Fund of the University of Amsterdam for their support to the science gateway actitivies at the AMC. 
	 We are also greatful to the financial support provided by FP7 E-INFRASTRUCTURE program for financial support to SCI_BUS, SHIWA and ER-flow projects.  
	 My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}

>>>>

\begin{abstract}
  \closednote{sil}{i find this a dry start. what about an intro sentence?
    Workflow engines are critical for the efficient and transparent
    exploitation of distributed infrastructures in the ecosystem of
    tools and services offered by science gateways.}{Tristan}{Done, thanks.}  Workflow engines
  are critical for the efficient and transparent exploitation of
  distributed infrastructures in the ecosystem of tools and services
  offered by science gateways.  We investigate six software
  architectures commonly used to integrate workflow engines into
  science gateways. In \emph{tight integration}, the workflow engine
  shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow import}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}.\closednote{sil}{what are the insights for wf engine developers?. this is
    not clear in the discussion and conclusions. i think the focus
    lies more on the integration and not on one or the other}{Tristan}{Fixed by your next note.} Tight
  integration and task encapsulation are the easiest to integrate and
  the most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be
    made more concise and poignant}{Tristan}{Added a description of
    the architectures, following @pbellec's suggestion.}
  \closednote{sil}{what about closing with this sentence from above: These
      results provide insights for science gateway architects and
      workflow engine developers. }{Tristan}{Done, thanks.}These results provide insights for
		science gateway architects and developers.
\end{abstract}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down to
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores, as explained in the Table caption. For each
criterion, a global score is computed by summing up the individual
scores for each metric. We ensure that the different metrics used to
calculate the global score for a criterion are comparable, so that
they can be combined meaningfully. The criteria and metrics are
explained hereafter.

\closednote{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}{Tristan}{Added a sentence about it.}

@@

\closednote{Sil}{i was wondering whether it would be more clear to present the metrics like this:
\item \texttt{I$_1$}: total number of components, and now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface}{Tristan}{Fixed, following discussion in https://github.com/glatard/FGCS-works-special-issue/pull/20}
Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. \closednote{Sil}{add here an example of an interaction
    that is not, for example, data transfers?}{Tristan}{Done.}
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
  \closednote{Si}{new interface = new interaction? maybe better to use
    one term only here. Maybe also refer to how this shows on figure
    2}{Tristan}{Fixed the interface vs. interaction issue.}
\item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated

@@

  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  \closednote{Silv}{it seems to me that this
    metric cannot be calculated in a straighforward manner from the
    diagrams.. is that correct? if it can, give an example of how this
    is seen in the diagram}{Tristan}{Done}
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. \closednote{sil}{again
    here it would be good to indicate how this is seen in the diagram}{Tristan}{done}

@@

\item Multiple engine instances (\texttt{S$_1$}): measures the potential \closednote{sil}{are we measuring possibility or potential? i think the second, because it is hypothetical. possibility brings me the idea of it is possible or not - which would be zero to almost all these features}{Tristan}{Fixed everywhere} of having

@@

  started. In this case, elastic engines can be implemented through some kind
  of auto-scaling mechanism to control the number

@@

  intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine

@@

\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing

@@

  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by

@@

\item Task scheduling (\texttt{S$_3$}): measures the complexity
  introduced by the architecture regarding task scheduling on
  the infrastructure. \closednote{sil}{need to be more specific about what
    is being called task scheduling here. also explain why it is
    difficult, or dont say anything if it is common knowledge}{Tristan}{OK, fixed.} Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not introduce any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.

@@

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
  \texttt{O$_1$}=1 otherwise.
  \note{si}{why here we do not have the "possible to implement" situation?}

@@

\paragraph{Scalability} The service architecture in principle supports
multiple engine instances through multiple workflow services. Adding a
new engine instance, however, requires manual intervention to declare
the new instance in the science gateway
(\texttt{S$_1$}=1). Consequently, elastic engines are difficult to
implement because they require a mechanism to update the science
gateway configuration when a new engine instance is
available. Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1).  \answerednote{sil}{is this score for
  VIP? i think that the score depends on the implementation of the
  science gateway, which could adopt some elastic mechanism within. so
  we should not say s_2=1 for the architecture}{Tristan}{Removed VIP example. Score definition still has to be modified according to the discussion in @SilviaOlabarriaga's PR (\#20).} The scheduling of
tasks on the infrastructure is as complex as in any other architecture
since the workflow engine might implement any kind of scheduling
policy (\texttt{S$_3$}=0).  \note{sil}{see comment in section 3.1}

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
\closednote{sil}{i have the impression that the comments below are about robustness, and not so much about scalability. i also think that the level of detail here is much larger than in other 3.X sections. finally, i wonder whether what is explained here (which i must admit i could not follow completely in the scope of this evaluation or architectures) is valid only for this architecture in particular. i have the impression that this argument is more general }{Tristan}{Rejected :) See discussion in PR #20.}
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. \closednote{Naj}{The lay reader wonders why?  an
  example like the one offered for interdependencies can
  help}{Tristan}{Done, thanks.} Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen. \closednote{Naj}{so is this the solution to
  fixing the above mentioned limitations with
  scalability?}{Tristan}{Unfortunately this is not a solution. We just
  say that this situation is unlikely to happen which is true.}

@@

\paragraph{Scalability} As in the service architecture, adding a new
workflow instance requires manual configuration in the science gateway
(instance of a parent engine), or in the parent engine (instance of a
child engine) (\texttt{S$_1$=1}). Similarly, elastic engines are
difficult to achieve. Distributed engines can be implemented through meta-workflows (\texttt{S$_2$=0}). Task scheduling
is more complex than in other architectures though, due to the fact
that workflow execution is split in different engines (\texttt{S$_3$=1}).

\paragraph{Specific features}
Meta-workflows can be implemented, which is one of the main interest of this
architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
science gateway cannot directly access fine-grained information in the
sub-workflow (\texttt{O$_2$=1}).

@@

Finally, workflow engines are only one of the many aspects involved in
the design of a science gateway. Other properties definitely influence
the design of a software architecture, for instance collaborative
features, data visualization, search, authentication, accounting and
so on.
\closednote{sil}{do you want to refer to the science gateway canvas paper here?}{Tristan}{Done (by you) in the related work section.}

\section{Related work}
\label{sec:related}

The early work~\cite{yuTaxonomy} proposes a generic architecture for
grid workflow systems that is based on the workflow reference model
defined by the Workflow Management Coalition~\cite{wfRefModel}.
Deelman et al. \cite{deelman2009workflows} have further characterized
the features of workflow systems and proposed a general architecture??
These two works, however, do not mention science gateways.

Shahand et al.~\cite{shahand:2015ab} presents the Science Gateway
Canvas, which is a business reference model where the most relevant
functional components are organized into functional groups. Although
workflow management is mentioned as a possible component to coordinate
distributed computations in science gateways, the paper does not
comment on architecture to achieve this.

Olabarriaga et al.~\cite{olabarriaga2014} presents a user-centered
view of the ecosystem of science gateways with focus on workflow
management. Their proposed layered system architecture includes
gateway, workflow engine and distributed infrastructure components,
similarly to the "service invocation" pattern. However other patterns
are not identified.

\section{Conclusion}

@@

This work has been made possible with the support of the Irving Ludmer
Family Foundation and the Ludmer Centre for Neuroinformatics and
Mental Health. The integration between PSOM and CBRAIN was supported
by a Brain Canada Platform Support Grant, as well as the Canadian
Consortium on Neurodegeneration in Aging (CCNA), through a grant from
the Canadian Institute of Health Research and funding from several
partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
LABX-0063) of Universit\'e de Lyon, within the program
``Investissements d’Avenir'' (ANR-11-IDEX-0007) operated by the French
National Research Agency (ANR). This work also falls into the scope of
the scientific topics of the French National Grid Institute (IdG). The
VIP team thanks the site administrators of the European Grid
Initiative and the GGUS support for their help related to the VIP
platform. The CBRAIN team is grateful for the computing cycles,
storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from
CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
national e-Infrastructure with the support of SURF Cooperative, the
Dutch national program COMMIT/ and the High Performance Computing and
Networking (HPCN) Fund of the University of Amsterdam for their
support to the science gateway actitivies at the AMC.  We are also
grateful to the financial support provided by FP7 E-INFRASTRUCTURE
program for financial support to SCI-BUS, SHIWA and ER-flow
projects. \closedanswerednote{Marc-e}{Would it be proper to thank EGI?
  Compute Canada? CANARIE?}{Tristan}{Added EGI and Compute Canada. Not
  sure if Canarie should be added. What about Calcul QC?}{Marc-e}{CQ,
  no. CANARIE, I'd say yes since they funded a major chunk of CBRAIN's
  development.} \closednote{sil}{should we then also thank the dutch crowd?
	 We also acknowledge the Dutch national e-Infrastructure with the support of SURF Cooperative,  the Dutch national program COMMIT/ and the High Performance Computing and Networking (HPCN) Fund of the University of Amsterdam for their support to the science gateway actitivies at the AMC. 
	 We are also greatful to the financial support provided by FP7 E-INFRASTRUCTURE program for financial support to SCI_BUS, SHIWA and ER-flow projects. My participation has been minimal, and the amc gateways are not discussed in detail (my fault), so maybe it would be an overkill}{Tristan}{done.}
